Nextstrain: real-time tracking of pathogen evolution

### 2018-04-17 whitepaper.tex
small edits

The regularly updated nature and rapidity of these analyses is crucial to the monitoring and understanding of pathogen epidemiology and evolution.
Sequencing times and costs are continually dropping, with on-the-ground sequencing used during recent epidemics \citep{quick2016real,faria2017epidemic}.
Rapid methods by which to analyse, interpret, and disseminate results must complement this sepeed of sequencing.

Nextstrain consists of data curation, analysis and visualisation components:
Python scripts maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI (www.ncbi.nlm.nih.gov), GISAID (www.gisaid.org) and ViPR (www.viprbrc.org), as well as GitHub repositories and other sources of genomic data.
A suite of tools perform phylodynamic analysis \citep{voltz2013}, including subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait geographic reconstruction, including inference of the most likely transmission events.
This leverages the maximum likelihood analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n$=1581 genomes) in under 2 hours on a modern laptop.
These scripts separate generic core functionality from a light pathogen-specific layer such that they are easily co-opted to different pathogens.
Visualisation of these analysis is available through \href{http://nextstrain.org}{nextstrain.org}.
This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or low mutation rate to confound phylogenetic signal.

@@

Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probable transmission events given the sampled data, together with a marginal likelihood for each event.
Internal node colours indicate inferred transmissions on the tree, which are also drawn as links between demes on the map.
It is important to make known the uncertainty of such reconstructions.
Confidence is conveyed by matching colour saturation to the confidence of that trait, and by displaying all relevant information when one hovers over the corresponding branch or isolate on the tree.
Sampling bias can obscure transmission links, and in certain cases we have chosen not to display the inferred states.

>>>>

The regularly updated nature and rapidity of these analyses is crucial to the monitoring and understanding of pathogen epidemiology and evolution.
Sequencing times and costs are continually dropping, with on-the-ground sequencing used during recent epidemics \citep{quick2016real,faria2017epidemic}.
Rapid methods by which to analyse, interpret, and disseminate results must complement this speed of sequencing.

Nextstrain consists of data curation, analysis and visualisation components:
Python scripts maintain a database of available sequences and related metadata, sourced from public repositories such as NCBI (www.ncbi.nlm.nih.gov), GISAID (www.gisaid.org) and ViPR (www.viprbrc.org), as well as GitHub repositories and other sources of genomic data.
A suite of tools perform phylodynamic analysis \citep{voltz2013}, including subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait geographic reconstruction, including inference of the most likely transmission events.
This leverages the maximum likelihood analyses implemented in TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n$=1581 genomes) in under 2 hours on a modern laptop.
These scripts separate generic core functionality from a light pathogen-specific layer such that they are easily adapted to different pathogens.
Visualisation of these analysis is available through \href{http://nextstrain.org}{nextstrain.org}.
This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and Nextstrain is able to be extended to most outbreaks with readily accessible genomic data, although we note the potential for recombination or low mutation rate to confound phylogenetic signal.

@@

Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probable transmission events given the sampled data, together with a marginal likelihood for each event.
Internal node colours indicate inferred transmissions on the tree, which are also drawn as links between demes on the map.
It is important to make known the uncertainty of such reconstructions.
Confidence is conveyed by matching colour saturation to the confidence of that trait, and by displaying all relevant information when one hovers over the corresponding branch or isolate on the tree.
Sampling bias and lack of data can obscure transmission links, and in certain cases we have chosen not to display the inferred states.
